# Practical Machine Learning Final Project
## Anirban Dhar

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to quantify how well they do it. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har
If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.


## Loading & Processing the Datasets
Load the Training & Test Dataset

```{r}
library(data.table, quietly=TRUE)
library(caret, quietly=TRUE)
library(randomForest, quietly=TRUE)
library(rpart, quietly=TRUE)
library(rpart.plot, quietly=TRUE)


dfTrain <- read.csv("pml-training.csv",na.strings = c('NA','#DIV/0!',''))

dfTest <- read.csv("pml-testing.csv",na.strings = c('NA','#DIV/0!',''))

```


# Split the Training Dataset into two parts

The test dataset will be split with 80% as  sample and 20% as cross validation data

```{r}
tempTrain <- createDataPartition(dfTrain$classe, p=0.6, list=FALSE)
tempTraining <- dfTrain[tempTrain, ]
tempTesting <- dfTrain[-tempTrain, ]
dim(tempTraining); dim(tempTesting)
```

# Process & clean the data

```{r}
nzv <- nearZeroVar(tempTraining, saveMetrics=TRUE)
tempTraining <- tempTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(tempTesting,saveMetrics=TRUE)
tempTesting <- tempTesting[,nzv$nzv==FALSE]
tempTraining <- tempTraining[c(-1)]
trainingV3 <- tempTraining
for(i in 1:length(tempTraining)) {
    if( sum( is.na( tempTraining[, i] ) ) /nrow(tempTraining) >= .7) {
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(tempTraining[i]), names(trainingV3)[j]) ) == 1)  {
                trainingV3 <- trainingV3[ , -j]
            }   
        } 
    }
}
tempTraining <- trainingV3
rm(trainingV3)
clean1 <- colnames(tempTraining)
clean2 <- colnames(tempTraining[, -58])  
tempTesting <- tempTesting[clean1]         
dfTest <- dfTest[clean2]             


dim(tempTesting)
dim(dfTest)
```
## Coerce datasets into same type

```{r}
for (i in 1:length(dfTest) ) {
    for(j in 1:length(tempTraining)) {
        if( length( grep(names(tempTraining[i]), names(dfTest)[j]) ) == 1)  {
            class(dfTest[j]) <- class(tempTraining[i])
        }      
    }      
}

dfTest <- rbind(tempTraining[2, -58], dfTest)
dfTest <- dfTest[-1,]
```


# Predict using Random Forests 
```{r}
set.seed(2000)
modFitRandFor <- randomForest(classe ~ ., data=tempTraining)
plot(modFitRandFor)

predictRandFor <- predict(modFitRandFor, tempTesting, type = "class")
confMatr <- confusionMatrix(predictRandFor, tempTesting$classe)
confMatr
plot(confMatr$table, col = confMatr$byClass, main = paste("Confusion Matrix With Random Forest: Accuracy =", round(confMatr$overall['Accuracy'], 5)))
```

# Predict using Decision Trees 
```{r}
set.seed(2000)
modFitDT <- rpart(classe ~ ., data=tempTraining, method="class")
rpart.plot(modFitDT)
predictionDT <- predict(modFitDT, tempTesting, type = "class")
confMatrDT <- confusionMatrix(predictionDT, tempTesting$classe)
confMatrDT
plot(confMatrDT$table, col = confMatrDT$byClass, main = paste("Confusion Matrix With Decision Tree: Accuracy =", round(confMatrDT$overall['Accuracy'], 5)))

```

# Course Project Prediction Quiz Results
As Random Forest model provided an accuracy of 99.89% compared to Decision Tree Model which provided an accuracy of 85.35%, using the Random Forest model for course quiz prediction

```{r}
predictCourseQuiz <- predict(modFitRandFor, dfTest, type = "class")
predictCourseQuiz
```

